Investigating Predictors for American Attitudes on Justifiability of Abortion

### 2022-06-20 Final/FinalReport.qmd
added captions and fixed errors

The dataset used in this analysis observes attitudes on the
justifiability of abortion among respondents in in the United States
across six "waves" of the World Values Survey (1982-2011), which is
administered every few years and collects information about people's
values and beliefs worldwide, alongside basic demographic
characteristics. The data are a nationally representative sample of the
American population and are collected via face-to-face interviews at the
respondents' homes. In this data set, the response variable,
justifiability of abortion, is a numerical measure on a scale of 1 to 10
on the individual person's attitude toward whether abortion is
justifiable or not. Individuals responded 1 for "abortion is never
justified" and 10 for "abortion is always justified.

@@

```{r visualization-summary}
#| warning: false
#| echo: false
ggplot(abortion_data, aes(x = aj)) +
  geom_histogram(binwidth = 1, aes(y=..density..), fill = "light blue") + 
  geom_density(color = "red") +
  labs(title = "Histogram of Abortion Attitudes",
       x = "Abortion Justifiability (1-Least, 10-Most)")
```

@@

title1 <- ggdraw() + 
  draw_label(
    "Distributions of Predictor variables",
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

top_row1 <- plot_grid(h1,h2,h3, ncol = 3, nrow = 1)
bottom_row1 <- plot_grid(h4,h5,h6, ncol = 3, nrow = 1)
plot_grid(title1, top_row1, bottom_row1, 
          ncol = 1, rel_heights = c(0.1, 1,1))
# ADD CAPTION
```

@@

plot_grid(b1, b2, b3, b4, b5, b6, ncol = 3, nrow = 2)

# ADD TITLE AND CAPTION
```

```{r correlations}
#| warning: false
#| echo: false
ggpairs(abortion_data, 
  columns = c("year", "ideology", "cai", "godimportant", "respectauthority", "nationalpride"),
  columnLabels = c("YS", "PI", "CAI", "IG", 
                   "RA", "NP")) +
   theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(angle = 45, size = 9),
    strip.text.y = element_text(angle = 0, hjust = 0)
    )
# ADD TITLE AND CAPTION
```

\pagebreak

@@

We conducted a 75% - 25% data split into training and testing sets,
using the random seed $206$.

@@

```{r recipe2}
#| echo: false
abortion_rec2 <- recipe(aj ~ year + godimportant,
                        data = abortion_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
```

```{r recipe3}
#| echo: false
abortion_rec3 <- recipe(aj ~ year + ideology + cai + nationalpride,
                        data = abortion_data) %>%
  step_center(ideology) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
```

```{r workflows}
#| echo: false
abortion_wflow <- workflow() %>%
  add_model(abortion_spec) %>%
  add_recipe(abortion_rec)

@@

```{r fit-models}
#| echo: false
abortion_fit <- abortion_wflow %>%
  fit(data = abortion_training)

@@

Fitting the MLR, we obtain the following table of parameter point
estimates.

```{r display-model}
#| echo: false
tidy(abortion_fit) %>%
  kable(digits = 3)
```

@@

```{r ci}
#| echo: false
df <- tidy(abortion_fit, conf.int = TRUE)
intercept <- df[1,]
PI_IG <- df[c(2,4),]
CAI <- df[3,]
YS <- df[c(5,6,7),]
RA_NP <- df[c(8,9,10),]

@@

top_row <- plot_grid(p2,p5, ncol = 2, nrow = 1)
bottom_row <- plot_grid(p1,p3,p4, ncol = 3, nrow = 1)
plot_grid(title_factor, top_row, title_numeric, bottom_row, 
          ncol = 1, rel_heights = c(0.1, 1,0.1,1))
#ADD CAPTION
```

@@

However, calculating the R-squared and root-mean-squared-error (RMSE)
values of our model shows that our model is not very strong at
predicting attitude toward the justifiability of abortion, with a R\^2
of 0.247 and RMSE of 2.53.

@@

To see how well our model fits the data on the attitudes toward the
justifiability of abortion, we conducted a 10-fold cross-validation,
calculating various metrics -- the Akaike information criterion (AIC),
Bayesian Information Criterion (BIC) and adjusted R-squared value -- to
evaluate the performance of our model on 10 subsets of our data.

@@

ggplot(metrics %>% filter(.metric == "rsq"),
	aes(x = id, y = .estimate, group = model, color = model)) +
    geom_point() +
    geom_line() +
    labs(title = "R-squared")
```

@@

final_table = as.table(abortion_table)
final_table %>%
  kable(digits = 3)
```

@@

Conditions check

Linearity \[residual plot vs fitted values\]

Randomness

Independence \[residual plot vs id\]

>>>>

The dataset used in this analysis observes attitudes on the
justifiability of abortion among respondents in the United States
across six "waves" of the World Values Survey (1982-2011), which is
administered every few years and collects information about people's
values and beliefs worldwide, alongside basic demographic
characteristics. The data are a nationally representative sample of the
American population and are collected via face-to-face interviews at the
respondents' homes. In this data set, the response variable,
justifiability of abortion, is a numerical measure on a scale of 1 to 10
on the individual person's attitude toward whether abortion is
justifiable or not. Individuals responded 1 for "abortion is never
justified" and 10 for "abortion is always justified.

@@

```{r vis-response}
#| warning: false
#| echo: false
ggplot(abortion_data, aes(x = aj)) +
  geom_histogram(binwidth = 1, aes(y=..density..), fill = "light blue") + 
  geom_density(color = "red") +
  labs(title = "Histogram of Abortion Attitudes",
       x = "Abortion Justifiability (1-Least, 10-Most)",
       caption = "Fig.1: The visual shows that the distribution of attitude toward abortion is trimodal, rather than bell-shaped. This may worsen our model's ability to fit the data.")+
  theme(plot.caption = element_text(hjust = 0.5))
```

@@

title1 <- ggdraw() + 
  draw_label(
    "Distributions of variables of interest",
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

caption1 <- ggdraw() + 
  draw_label(
    "Fig.2: The visuals show that while some variables in the data set have bell-shaped distributions, others are left-skewed.",
    x = 0,
    hjust = 0, 
    size = 8
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

top_row1 <- plot_grid(h1,h2,h3, ncol = 3, nrow = 1)
bottom_row1 <- plot_grid(h4,h5,h6, ncol = 3, nrow = 1)
plot_grid(title1, top_row1, bottom_row1, caption1, 
          ncol = 1, rel_heights = c(0.1, 1,1, 0.1)) 
```

@@

title2 <-  ggdraw() + 
  draw_label("Boxplots of Abortion Attitudes against interesting variables",
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

caption2 <- ggdraw() + 
  draw_label("Fig.3: The visuals show that there may be a relationship between the response and each of these variables, making them likely predictors.",
    x = 0,
    hjust = 0, size = 7
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

middle <- plot_grid(b1, b2, b3, b4, b5, b6, ncol = 3, nrow = 2)

plot_grid(title2, middle, caption2, 
          ncol = 1, rel_heights = c(0.1, 1, 0.1))

```{r correlations}
#| warning: false
#| echo: false
ggpairs(abortion_data, 
  columns = c("year", "ideology", "cai", "godimportant", "respectauthority", "nationalpride"),
  columnLabels = c("YS", "PI", "CAI", "IG", 
                   "RA", "NP")) +
   theme(
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(angle = 45, size = 9),
    strip.text.y = element_text(angle = 0, hjust = 0),
    plot.caption = element_text(hjust = 0.5)
    ) +
  labs(title = "Correlation Matrix of Predictors of Interest", caption = "Fig.4: This matrix shows that the correlations between pairs of predictors are low, with the highest being 0.370.")

```
\pagebreak

@@

We conducted a 75% - 25% data split into training and testing sets,
using the random seed $206$. Fitting the MLR, we obtain the following table of parameter point estimates.

@@

```{r recipe2}
#| echo: false
#| warning: false
abortion_rec2 <- recipe(aj ~ year + godimportant,
                        data = abortion_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
```

```{r recipe3}
#| echo: false
#| warning: false
abortion_rec3 <- recipe(aj ~ year + ideology + cai + nationalpride,
                        data = abortion_data) %>%
  step_center(ideology) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())
```

```{r workflows}
#| echo: false
#| warning: false
abortion_wflow <- workflow() %>%
  add_model(abortion_spec) %>%
  add_recipe(abortion_rec)

@@

```{r fit-models}
#| echo: false
#| warning: false
abortion_fit <- abortion_wflow %>%
  fit(data = abortion_training)

@@

```{r display-model}
#| echo: false
tidy(abortion_fit) %>%
  kable(digits = 3, caption = "This table displays the estimate, standard error, statistic and p.value for each parameter of the model.")
```

@@

```{r ci}
#| echo: false
#| warning: false
df <- tidy(abortion_fit, conf.int = TRUE)
intercept <- df[1,]
PI_IG <- df[c(2,4),]
CAI <- df[3,]
YS <- df[c(5,6,7),]
RA_NP <- df[c(8,9,10),]

@@

top_row <- plot_grid(p2,p5, ncol = 2, nrow = 1)
bottom_row <- plot_grid(p1,p3,p4, ncol = 3, nrow = 1)
caption3 <- ggdraw() + 
  draw_label("Fig.5: 95% confidence intervals of estimates for model parameters, using points and error bars.",
    x = 0,
    hjust = 0, size = 10
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

middle <- plot_grid(b1, b2, b3, b4, b5, b6, ncol = 3, nrow = 2)

plot_grid(title_factor, top_row, title_numeric, bottom_row, caption3, 
          ncol = 1, rel_heights = c(0.1, 1,0.1,1,0.1))

@@

However, calculating the R-squared and root-mean-squared-error (RMSE)
values of our model shows that our model is not very strong at
predicting attitude toward the justifiability of abortion, with a R-squared
of 0.247 and RMSE of 2.53.

@@

To see how well our model fits the data on the attitudes toward the
justifiability of abortion, we conducted a 10-fold cross-validation with seed 206,
calculating various metrics -- the Akaike information criterion (AIC),
Bayesian Information Criterion (BIC) and adjusted R-squared value -- to
evaluate the performance of our model on 10 subsets of our data.

@@

ggplot(metrics %>% filter(.metric == "rsq"),
	aes(x = id, y = .estimate, group = model, color = model)) +
    geom_point() +
    geom_line() +
    labs(title = "R-squared", caption = "Fig.6: This plot of R-squared values demonstrates our chosen model 1 is the best of the three.")
```

@@

final_table = as.table(abortion_table)
final_table %>%
  kable(digits = 3, caption = "With the lowest AIC and BIC and highest adjusted R-squared value, our chosen model is best of the three.")
```

@@

Below we will check each of the following conditions for our model:
Linearity, Constant Variance, Normality, and Independence.

```{r fit-aug}
#| echo: false
final_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(aj ~ year + ideology + cai + godimportant +
                        respectauthority + nationalpride,
      data = abortion_data)

final_fit_aug <- augment(final_fit$fit)
```

As shown in the following plot of residuals against fitted values, the linearity condition is not met as the residuals are not scattered around the horizontal red line. The constant variance condition is also not met.

```{r linearity}
#| echo: false
c1 <- ggplot(data = final_fit_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Plot of residuals v. fitted values", x = "Predicted values", y = "Residuals")

c2 <- ggplot(final_fit_aug, aes(.resid)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1) +
  stat_function(fun = dnorm, 
    args = list(mean = mean(final_fit_aug$.resid), sd = sd(final_fit_aug$.resid)), 
    lwd = 2,  color = "red"  )+
 labs(title = "Distribution of residuals")

c3 <- ggplot(final_fit_aug, aes(y = .resid, x = 1:nrow(final_fit_aug))) +
  geom_point(alpha = 0.5) +
  labs(x = "Order of data collection", y = "Residuals", title = "Plot of residuals v. order of collection")

top_row1 <- plot_grid(c1,c2, ncol = 2, nrow = 1)

caption4 <- ggdraw() + 
  draw_label("Fig.7: The linearity condition and constant variance condition is clearly not met in this plot of residuals v. fitted values.",
    x = 0,
    hjust = 0, size = 8
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )
caption5 <- ggdraw() + 
  draw_label("Fig.8: The residuals appear to follow a normal distribution, which looks like a bell-shape, as illustrated in red.",
    x = 0,
    hjust = 0, size = 8
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )
caption6 <- ggdraw() + 
  draw_label("Fig.9: The residuals plotted by observation order are randomly scattered, showing no structure.",
    x = 0,
    hjust = 0, size = 8
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 7)
  )

plot_grid(top_row1, c3, caption4, caption5, caption6,
          ncol = 1, rel_heights = c(1, 1,0.1,0.1,0.1))
